Molecular Gas Dynamics of the lensed wet-merger RXSJ1131-1231 at $z$=0.654

### 2016-06-28 WRITEUP/ms.tex
done merger stage

In the rest-frame,
emission in the observed $H$-band corresponds to NIR emission,
tracing radiation from the accretion disk surrounding
the central AGN and also from old and evolved stellar populations;
$I$-band corresponds to roughly the optical $V$-band, tracing stellar radiation from
existing, less massive (\ie longer-lasting) stars;
$V$-band corresponds to roughly $U$-band,  tracing radiation from massive young stars
in the host galaxy. Hence,
the $V$-band compactness may be explained in part
due to the fact that its emission is
more susceptible to dust extinction than in other bands and/or
a central starburst caused by higher
concentrations of star-forming gas towards the central regions -- owing to
gravitational perturbations induced
from interactions with the companion
\citep[\eg][]{DiMatteo05a}.
This is consistent with the picture that old stars form first and constitute the bulge component
of a spiral galaxy and that nuclear starbursts (in the inner few kpc) are triggered in mergers.

@@

The SFE (using the above definition) is highly dependent on the gas mass.
Yet, the use of an ``ULIRG" \alphaco value is not well justified and 
should only be considered as a lower limit \citepalias[see][]{Solomon97a}.
It may be more appropriate to use an \alphaco closer to the ``Galactic" value in 
some ULIRGs \citep[\eg][]{Papadopoulos12a}. 
From numerical models, it is found that mergers on average have lower \alphaco values than
disk galaxies \citep{Narayanan12a}.
But since the gas temperatures and velocity dispersions in minor mergers do not deviate from 
galactic GMCs as much as in major mergers, the \alphaco may not be as extreme as 0.8\,\alphaU.
% thus the rise in CO intensity is not as large, and the X-factor (and thus alphaco) does not decrease as much {Narayana11a}.
Hence,
we can derive an upper limit on the gas mass for RXJ1131 using the standard conversion factor.
If we have instead applied the Galactic value of 4.6\,\alphaU, 
the gas mass (and thus the SFE) would be higher by a factor of $\sim$6. 
In contrast to results in the classical study of nearby ULIRGs by \citetalias{Downes98a}, who find that
the gas mass exceeds the dynamical mass using 
the Galactic conversion factor 
(and thus inappropriate for their sample), the gas mass derived for
RXJ1131 using the Galactic value is consistent with its dynamical mass.
%Obviously, this will lead to a longer depletion time.

@@

%--------------------------------------------------------------------------
%                                Discussion
%--------------------------------------------------------------------------
\section{Discussion} \label{sec:diss}
\subsection{Velocity Offset and Recoiling BH}
Using the redshift deduced from the CO line 
as the systemic redshift, we find
a velocity offset of $\sim$780\,\kms with the optical lines:
the broad Mg{\scriptsize II}~2798\AA\ and the narrow 
$[$O{\scriptsize III}$]$~4959, 5007\AA\ lines \citep[$z_{\rm AGN} = 0.658$;][]{Sluse03a}.
This indicates that the broad line region (BLR) of the AGN is at a
slightly higher redshift than the host galaxy.
From the CO renzogram and channel maps in \Fig{CO21mom} and \Fig{chanmap}, the emission 
corresponding to the the line center are not co-spatial
with the optical quasar. In particular, the quasar emission forms 
the point-like images along the lensing arc, which is 
spatially offset to the NW
of the CO line center emission.
Spatial offsets between optical BLRs and CO emission are not uncommon
but such a high kinematic offset is not commonly observed.
For instance, the
velocity offset between AGNs and their hosts galaxies in 
an SDSS sample of ongoing mergers at $z$\,$<$\,0.21 is at most $\sim$410\,km/s \citep{Comerford14a}.
The same authors also suggest mechanisms
such as AGN outflows, jets, and gravitational recoil of SMBHs in post-merger
as plausible explanations for these offsets.

@@

In fact, a large velocity offset can also result from a recoiling BH where its
BLR is moving at high velocity relative to the bulk of the host galaxy and
the ionized gas \citep{Madau04a, Bonning07a, Loeb07a}. % narrow AGN lines.
Such a scenario may occur after galaxies and their BHs merge, during which 
their orbital energy is being released as gravitational wave to overcome the last stable orbit.
This results in a recoil in the merged BH for an uneven BH mass merger as a
non-zero net angular momentum is also being carried away.
Depending on their initial conditions, % mass ratios, spin, rotational direction, 
numerical relativity simulation have shown that
the recoil velocity can reach up to $\sim$4000\,\kms for spinning BHs \citep[\eg][]{Campanelli07a}.
% The BH is displaced from the galactic nucleus in this scenario, and might 
% return to the center in a few dynamical time-scales 
% by dynamical friction or be kicked out of the gravitational potential of the host galaxy.
This together with the fact that the BH in RXJ1131 has a high spin parameter \citep{Reis14a} 
renders the recoiling BH scenario a viable option for the origin
of the optical velocity offset.
To reconcile this picture with the fact that a companion is still observed, 
the system may be at its subsequent minor merger stages but was intially
a major merger. 
This would imply previous encounters between RXJ1131 and its companion, 
which would also explain the spinning BH.

@@

%==============================================================================
%                                Back matters
%==============================================================================

% ACKNOWLEDGEMENTS
%-----------------
\acknowledgments
This work is based on observations carried out under project number S14BX001
with the IRAM NOEMA Interferometer. IRAM is supported by INSU/CNRS (France), MPG (Germany) and IGN (Spain).
Support for CARMA construction was derived from the Gordon and Betty Moore
Foundation, the Kenneth T. and Eileen L. Norris Foundation, the James S.
McDonnell Foundation, the Associates of the California Institute of
Technology, the University of Chicago, the states of Illinois, California, and
Maryland, and the National Science Foundation. Ongoing CARMA development and
operations are supported by the National Science Foundation under a
cooperative agreement and by the CARMA consortium universities.
The National Radio Astronomy Observatory is a facility of the National Science
Foundation operated under cooperative agreement by Associated
Universities, Inc.
This research made use of data obtained with {\it Herschel}, an ESA space
observatory with science instruments provided by European-led Principal
Investigator consortia and with important participation from NASA.
This research has made use of NASA's Astrophysics Data System Bibliographic
Services.
This work is based in part on observations made with the \spitzer,
which is operated by the Jet Propulsion Laboratory, California Institute of
Technology under a contract with NASA.
This publication made use of data products from the Wide-field Infrared
Survey Explorer, which is a joint project of the University of California, Los
Angeles, and the Jet Propulsion Laboratory/California Institute of Technology,
funded by the National Aeronautics and Space Administration.
This publication made use of data products from the Two Micron All Sky
Survey, which is a joint project of the University of Massachusetts and the
Infrared Processing and Analysis Center/California Institute of Technology,
funded by the National Aeronautics and Space Administration and the National
Science Foundation.
This research made use of the NASA/IPAC Extragalactic Database (NED) which
is operated by the Jet Propulsion Laboratory, California Institute of
Technology, under contract with the National Aeronautics and Space
Administration.
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy}.
This research made use of APLpy, an open-source plotting package for Python hosted at \url{http://aplpy.github.com}.

>>>>

In the rest-frame,
emission in the observed $H$-band corresponds to NIR emission,
tracing radiation from the accretion disk surrounding
the central AGN and also from old and evolved stellar populations;
$I$-band corresponds to roughly the optical $V$-band, tracing stellar radiation from
existing, less massive (\ie longer-lasting) stars;
$V$-band corresponds to roughly $U$-band,  tracing radiation from massive young stars
in the host galaxy. Hence,
the $V$-band compactness may be explained in part
due to the fact that its emission is
more susceptible to dust extinction than in other bands and/or
a central starburst caused by higher
concentrations of star-forming gas towards the central regions -- owing to
gravitational perturbations induced
from interactions with the companion
\citep[\eg][]{DiMatteo05a}.
This is consistent with the picture that old stars form first and constitute the bulge component
of a spiral galaxy and that nuclear starbursts (in the inner few kpc) can be triggered latter in mergers.

@@

The SFE (using the above definition) is highly dependent on the gas mass.
Yet, the use of an ``ULIRG" \alphaco value is not well justified and 
should only be considered as a lower limit \citepalias[see][]{Solomon97a}.
It may be more appropriate to use an \alphaco closer to the ``Galactic" value in 
some ULIRGs \citep[\eg][]{Papadopoulos12a}. 
From numerical models, it is found that mergers on average have lower \alphaco values than
disk galaxies \citep{Narayanan12a}.
But since the gas temperatures and velocity dispersions in minor mergers do not deviate from 
galactic GMCs as much as in major mergers, the \alphaco may not be as extreme as 0.8\,\alphaU.
% thus the rise in CO intensity is not as large, and the X-factor (and thus alphaco) does not decrease as much {Narayana11a}.
Hence,
we can derive an upper limit on the gas mass for RXJ1131 
using the Galactic value of 4.6\,\alphaU. This yields
a gas mass (and thus a SFE) that is $\sim$6 times higher.  
In contrast to results in the classical study of nearby ULIRGs by \citetalias{Downes98a}, who find that
the gas mass exceeds the dynamical mass using 
the Galactic conversion factor 
(and thus inappropriate for their sample), the gas mass derived for
RXJ1131 using the Galactic value is consistent with its dynamical mass.
%Obviously, this will lead to a longer depletion time.

@@

%--------------------------------------------------------------------------
%                                Discussion
%--------------------------------------------------------------------------
\section{Discussion} \label{sec:diss}
\subsection{Velocity Offset and a Recoiling Black Hole}
Using the redshift deduced from the CO line 
as the systemic redshift, we find
a velocity offset of $\sim$780\,\kms with the optical lines:
the broad Mg{\scriptsize II}~2798\AA\ and the narrow 
$[$O{\scriptsize III}$]$~4959, 5007\AA\ lines \citep[$z_{\rm AGN} = 0.658$;][]{Sluse03a}.
This indicates that the broad line region (BLR) of the AGN is at a
slightly higher redshift than the host galaxy.
From the CO renzogram and channel maps in \Fig{CO21mom} and \Fig{chanmap}, the emission 
corresponding to the the line center are not co-spatial
with the optical quasar. In particular, the quasar emission forms 
the point-like images along the lensing arc, which is 
spatially offset to the NW
of the CO line center emission.
Spatial offsets between optical BLRs and CO emission are not uncommon
but such a high kinematic offset is not commonly observed.
For instance, the
velocity offset between AGNs and their hosts galaxies in 
an SDSS sample of ongoing mergers at $z$\,$<$\,0.21 is at most $\sim$410\,km/s \citep{Comerford14a}.
The same authors also suggest mechanisms
such as AGN outflows, jets, and gravitational recoil of SMBHs in post-merger
as plausible explanations for these offsets.

@@

In fact, a large velocity offset can also result from a recoiling black hole (BH) where its
BLR is moving at high velocity relative to the bulk of the host galaxy and
the ionized gas \citep{Madau04a, Bonning07a, Loeb07a}. % narrow AGN lines.
Such a scenario may occur after galaxies and their BHs merge, during which 
their orbital energy is being released as gravitational wave to overcome the last stable orbit.
This results in a recoil in the merged BH for an uneven BH mass merger as a
non-zero net angular momentum is also being carried away.
Depending on their initial conditions, % mass ratios, spin, rotational direction, 
numerical relativity simulations have shown that
the recoil velocity can reach up to $\sim$4000\,\kms for spinning BHs \citep[\eg][]{Campanelli07a}.
% The BH is displaced from the galactic nucleus in this scenario, and might 
% return to the center in a few dynamical time-scales 
% by dynamical friction or be kicked out of the gravitational potential of the host galaxy.
This together with the fact that the BH in RXJ1131 has a high spin parameter \citep{Reis14a} 
renders the recoiling BH scenario a viable option for the origin
of the optical velocity offset.
To reconcile this picture with the fact that a companion is still observed, 
the system may be in its subsequent stages of (minor) merging but was initially
a major merger. 
This would imply previous encounters between RXJ1131 and its companion, 
which would also explain the spinning BH.

@@

\subsection{Merger stage RXJ1131-1231}
%It is clear that the signature of a galaxy merger changes as the merging progresses.
%For instance, extended molecular disks are often associated with advanced mergers 
%\citep[\citetalias{Downes98a};][]{Gao99a}.but see also U14 for outliers
% with extended molecular gas in merger remnants

The (gas) mass ratio, the intrinsic CO line profile,  
% since the bulk of the angular momentum can be retained only in minor mergers if small separation
and the source-plane velocity field are all evident of 
a minor merger, in good agreement with an 
earlier study by \citetalias{Brewer08a} who
independently conclude the presence of a dwarf companion 
of size $\sim$700\,kpc across. Since minor mergers may in some cases
result from previous passage of major mergers, it is possible that RXJ1131
has encountered with the progenitor galaxy of the companion and is undergoing 
the subsequent series of minor merging.
One piece of evidence comes from % that they have encountered
the compact UV emission, which can be starburst resulting from gas accumulation 
owing to a non-axisymmetric perturbation from the companion.
% the stellar component torques away the angular momentum, allowing the gas to fall inward.
The highly spinning black hole in RXJ1131 is also consistent with 
this hypothesis. 
If we favor the gravitational wave recoil mechanism 
to explain the optical velocity offset, it would also imply the black 
holes (and their host galaxies) have merged or are at their final coalescence. 
In addition, the source-plane CO position of the companion appears to be consistent with 
a picture where its molecular gas is mixed with RXJ1131 (but with large uncertainties). 
%\ie companion CO location along/within the RXJ1131 CO disc
% as well as one where its emission is co-spatial with the stellar light. 
The IR luminosity also agrees with 
the results from previous studies: AGNs are
typically found in late stage mergers 
\citep{Yuan10a,Iwasawa11a,Carpineti12a} and
only major mergers near their final
coalescence can provide \LIR$>$\,10$^{12}$\,\Lsun,
\citep[\eg][]{Carpineti15a,Larson16a}.

It may be puzzling as to why the stellar emission in the companion 
is spatially offset from RXJ1131 while its CO gas appears to be mixed-in if they 
have encountered.
This phenomenon is in fact not unique to our source. 
The gas cores of the individual 
% merging galaxies 
mergers NGC~6090 and NGC~6240 are also observed
to be coalescing  
% is in the process of or has already coalesced into a single component
while their stellar cores remain % distinct
separated \citep[0.5$-$3\,kpc;][]{Bryant99a}. 
Although this is difficult to reconcile with the fact that the stellar light in the companion 
originates from recent \SF (UV emission). 
More sensitive and higher resolution observations will be needed
to confirm any offsets between the CO gas and the stellar light in the companion.
% allow for a more detailed lens modeling to trace the gas distribution and morphology in the system
Another puzzle may be the absence of morphological features.
% tidal tails if they have already encountered.
One straightforward explanation may simply be the lack of spatial resolution.
Alternatively, it may be a result of the offset in timing between observable
morphological disturbances and the merger-induced starbursts \citep{Lotz08a}.
% with strong morphological disturbances occurring before bursts of merger-induced star formation and for shorter durations

\subsection{Fate of RXJ1131}
A classical picture of mergers is that they are responsible for the formation of local spheroids.
The growing consensus that more realistic
treatments of star formation and supernovae feedback ? which suppress
efficient early star formation ? play a critical role in enabling
disc formation in simulations.
Mergers are expected to evolve into
local ellipticals or large-disks (bulge dominated; Springel05a, Robertson06a)
and disc with low B/D ratio (Hopkins09a) from simulations.
Reason for suppression of bulge formation in mergers of gas-rich discs (even in major merger):
 because in systems that are sufficiently gas-rich, there's little/no significant (collisionless) stellar material available to torque away the cold gas in the disc; little/no angular momentum is lost, and gas will rapidly reform a cold disc in the background of the relaxing stellar potential (collisionless):
Gas, being collisional, cannot violently relax, but must
have its angular momentum torqued away in order to dissipate and
build a bulge by forming stars in a central starburst (\ie gas lose angular momnentum to stars in the perturbed disk and fall toward center). In a merger,
this torque is primarily internal, from stars in the same disc: the
passages and merger of the secondary induce a non-axisymmetric
distortion in the primary stellar and gas disc. The stellar distortion
(in a trailing resonance and close proximity to the gas response)
efficiently removes angular momentum from the gas and allows it
to fall to the center and form a bulge (see also Barnes \& Hernquist
1996; cold gas with no angular momentum will be largely unable to form any sort of disk, and that a galaxy?s worth of gas compressed to high densities and small radii will inevitably form a large mass in stars). But it no significant stellar feature is there (e.g. in the more gas-rich ones (Hopkins09a), the final product will have a bulge mass fraction $\sim$0.1$-$0.2).
i.e. small bulge disc can also form from minor merger or major merger (if gas fraction is sufficiently large)

% fate
But since the two galaxies are found to be gas-rich, can dissipates energy
through e.g. frictional forces and drive gas toward the center, leading to
higher rotational velocity and dispersion in the central region.
In that case,
it is unlikely that RXJ1131 will evolve into present-day massive ellipticals (M$\gtrsim$1\E{12}\,\Msun), since RXJ1131 is now a minor merger.
Instead, it is likely to evolve into ``normal'' ellipticals (depending on the feedback, since a major gas-rich merger without AGN feedback does not, by itself, produce a red and dead spheroidal galaxy (Lotz08a).) or massive disk galaxies where the stellar populations from the companion is dragged and added onto the dominant galaxy RXJ1131 to thicken the disc, consistent with recent models (disk from gas-rich major merger:
\citep[\eg][]{Lotz08a, Robertson06a}.

% evolve into disk
Given the
massive gas content distributed in an extended configuration/disc with its SFE on the low end of other ULIRGs, we suggest RXJ1131 may evolve into the
latter (large-disk) unless the ``companion'' somehow triggers some other
mechanisms in the latter stage, transporting the gas 
toward the central nuclear region and to simulate a final burst to form a large bulge or somehow destroy the
disc in RXJ1131.

%==============================================================================
%                                Back matters
%==============================================================================
% ACKNOWLEDGEMENTS
%-----------------
\acknowledgments
This work is based on observations carried out under project number S14BX001
with the IRAM NOEMA Interferometer. IRAM is supported by INSU/CNRS (France), MPG (Germany) and IGN (Spain).
Support for CARMA construction was derived from the Gordon and Betty Moore
Foundation, the Kenneth T. and Eileen L. Norris Foundation, the James S.
McDonnell Foundation, the Associates of the California Institute of
Technology, the University of Chicago, the states of Illinois, California, and
Maryland, and the National Science Foundation. Ongoing CARMA development and
operations are supported by the National Science Foundation under a
cooperative agreement and by the CARMA consortium universities.
The National Radio Astronomy Observatory is a facility of the National Science
Foundation operated under cooperative agreement by Associated
Universities, Inc.
This research made use of data obtained with {\it Herschel}, an ESA space
observatory with science instruments provided by European-led Principal
Investigator consortia and with important participation from NASA.
This research has made use of NASA's Astrophysics Data System Bibliographic
Services.
This work is based in part on observations made with the \spitzer,
which is operated by the Jet Propulsion Laboratory, California Institute of
Technology under a contract with NASA.
This publication made use of data products from the Wide-field Infrared
Survey Explorer, which is a joint project of the University of California, Los
Angeles, and the Jet Propulsion Laboratory/California Institute of Technology,
funded by the National Aeronautics and Space Administration.
This publication made use of data products from the Two Micron All Sky
Survey, which is a joint project of the University of Massachusetts and the
Infrared Processing and Analysis Center/California Institute of Technology,
funded by the National Aeronautics and Space Administration and the National
Science Foundation.
This research made use of the NASA/IPAC Extragalactic Database (NED) which
is operated by the Jet Propulsion Laboratory, California Institute of
Technology, under contract with the National Aeronautics and Space
Administration.
This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy}.
This research made use of APLpy, an open-source plotting package for Python hosted at \url{http://aplpy.github.com}.
